---
title: "ETC3580: Advanced Statistical Modelling"
author: "Week 6: Bootstrap"
fontsize: 14pt
output:
  beamer_presentation:
    theme: metropolis
    fig_height: 4.5
    fig_width: 7
    highlight: tango
    includes:
      in_header: header.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache=FALSE, warning=FALSE, message=FALSE,
  dev.args=list(bg=grey(0.9), pointsize=11))
library(faraway)
library(tidyverse)
library(forecast)
```

## Pull yourself up by your bootstraps

\fullwidth{Bootstraps}

## What is the bootstrap?

  * The bootstrap is a flexible statistical tool to **quantify the uncertainty** associated with a *given estimator*.
  * We **mimic the process of obtaining new data sets** of the same size, so that we can estimate the variability of our estimate.

    1. Repeatedly sample observations **from the original data set with replacement** (nonparametric)
    2. Repeatedly sample observations **from an estimated model** (parametric).

## Illustration of the nonparametric bootstrap

\fullheight{5-11}

## Bootstrap of sample median

 1. Draw a random sample of data with replacement.
 2. Calculate sample median of sample.

###
Repeat many times. Find percentiles of sample medians.

## Bootstrap of sample median

```{r boxplotincome, fig.height=1}
jbox <- function(y) {
  p <- ggplot() +
    geom_jitter(aes(x=1, y=y), alpha=0.1, col='blue',
                position=position_jitter(w=0.2,h=0))  +
    geom_boxplot(aes(x=1,y=y),
                 outlier.shape=NA) +
    coord_flip() + xlab("") + ylim(20,255) +
    theme(axis.line.y = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          panel.grid.major.y = element_blank(),
          panel.grid.minor.y = element_blank())
  return(p)
}
y <- fpp::credit$income
jbox(y) + ylab("Income")
```

\pause

```{r, fig.height=1}
jbox(sample(y,replace=TRUE)) + ylab("Bootstrapped data")
```

\pause

```{r, fig.height=1}
jbox(sample(y,replace=TRUE)) + ylab("Bootstrapped data")
```

\pause

```{r, fig.height=1}
jbox(sample(y,replace=TRUE)) + ylab("Bootstrapped data")
```

\pause

```{r, fig.height=1}
jbox(sample(y,replace=TRUE)) + ylab("Bootstrapped data")
```

## Nonparametric bootstrap of sample median

 * Percentiles of bootstrapped medians provide a confidence interval for the true median.

\fontsize{12}{14}\sf

```{r bootstrap1, echo=TRUE}
median(y)
bmedians <- numeric(100)
for(i in seq_along(bmedians))
  bmedians[i] <- median(sample(y, replace=TRUE))
quantile(bmedians, prob=c(0.025, 0.975))
```

## Parametric bootstrap of sample median

 * Assume a parametric model/distribution for data, and re-sample from the model.

```{r, fig.height=3.8}
gghistogram(y, add.normal=TRUE)
```

## Parametric bootstrap of sample median

 * Assume a parametric model/distribution for data, and re-sample from the model.
 * Take log of data.

```{r, fig.height=3.8}
gghistogram(log(y), add.normal=TRUE)
```

## Parametric bootstrap of sample median

 * Take log of data.

\fontsize{12}{14}\sf

```{r bootstrap2, echo=TRUE}
logy <- log(y)
mu <- mean(logy)
sigma <- sd(logy)
n <- length(y)
for(i in seq_along(bmedians))
  bmedians[i] <- exp(median(rnorm(n, mu, sigma)))
quantile(bmedians, prob=c(0.025, 0.975))
```

\pause

 * logNormal assumption not working well!
